A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2017-01-06 model.tex
tiny edits

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations (overstability). %Writing $\sigma =
%-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
%growth rates, respectively, 
The imaginary part of Eq. \ref{int_rel}
gives  
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}. \label{thermal_instability}
\end{align} 
This represents instability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag instabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/print-chap-full.pdf}}.

@@

\subsubsection{Physical requirement of dust-gas drag instabilities}

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
%  s = \frac{\imag\int P
%  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd 
%    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check} 
  s = \frac{\imag\int P
    \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\mathcal{I}^2}. \label{vsi_check} 
\end{align} 
This is VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We present 
numerical solutions of the VSI with perfectly-coupled dust in \S\ref{results}.

>>>>

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. %Writing $\sigma =
%-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
%growth rates, respectively, 
The imaginary part of Eq. \ref{int_rel}
gives  
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

It is well-known that such correlations can lead to pulsational
instabilities in stars \citep{cox67}. We thus interpret  
dust-drag overstabilities in a similar way, 
adapting from the treatment of stellar 
pulsations by \cite{cox67} and lecture notes by \cite{samadi15} and 
J. Christensen-Dalsgaard\footnote{\url{http://astro.phys.au.dk/$\sim$jcd/oscilnotes/print-chap-full.pdf}}.

@@

\subsubsection{Physical requirement of dust-gas drag overstabilities}

@@

\subsection{Locally isothermal gas perfectly coupled to dust}\label{dusty_vsi_int}
If $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel} 
gives  
\begin{align}
%  s = \frac{\imag\int P
%  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd 
%    v_r|^2 + |\dd v_z|^2\right)dV}. \label{vsi_check} 
  s = \frac{\imag\int P
    \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\mathcal{I}^2}, \label{vsi_check} 
\end{align} 
again assuming $\omega\neq0$. 
This is VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}. We present 
numerical solutions of the VSI with perfectly-coupled dust in \S\ref{results}.

### 2017-01-06 summary.tex
minor edits on SGI stuff
tiny edits

\subsubsection{Gravitational instability} %vertical structure 
The addition of dust enhances gravitational
instability (GI). This is because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\epsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the effective temperature may 
be lowered to enable instability.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of \emph{all}
instabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

>>>>

\subsubsection{Classic gravitational instabilities} %vertical structure 
The addition of dust enhances gravitational
instability (GI). This is because dust particles contribute to the
total disk mass but not thermal pressure, which effectively lowers the
disk temperature \citep[][]{thompson88,shi13}. For typical dust-loading 
$\epsilon, Z\ll1$, this effect is unimportant. However, if $\epsilon$ is
large (e.g. due to dust settling) then the effective temperature may 
be lowered to enable instability.

@@

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of 
overstabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies there exists a time interval where 
the gas pressure of a parcel of the mixture is increasing whilst dust is already being expelled.  
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.
